CODECHECK: An open-science initiative to facilitate sharing of computer programs and results presented in scientific publications.

### 2021-01-05 codecheck-paper.tex
add possible source
adding a couple of old notes (forgot to commit)

\begin{itemize}
\item Check with Giuliano re: ``code works''. Write to Naomi.
\item Use "execution" insteand of "sharing" in title?
\item Can we introduce "research compendium" as a term?
\item Should we make code and data sharing a prerequisite or leave that door open?
\item Take a look at the Lancet's statistical review process (areas for improvement by authors) \url{https://www.thelancet.com/journals/lancet/article/PII0140-6736(92)90409-V/fulltext}
\item How to relate to "reproducibility does not matter but trust does", see \url{https://gmbecker.github.io/MayInstituteKeynote2019/outline.html}
\item Should we ask the people currently listed in the acknowledgements?
\item revisit \url{https://github.com/codecheckers/discussion}
\item \url{https://hdsr.mitpress.mit.edu/pub/f0obb31j/release/1} and other articles form the HDSR special issue
\end{itemize}

@@

The time for a CODECHECK greatly depends on the intentions and the ascribed
importance. The earlier a CODECHECK happens in the process, the more the
results can play a role in the most important decision in scholarly 
communication: is a paper published, sent back for revisions, or rejected.
The later in a review process the check happens, the less challenging is it
to allow a bidirectional communication between author and codechecker, e.g.,
because the author might already be notified of the acceptance and may more
freely share materials online.
Even earlier checks, i.e., a CODECHECK of a preprint, may even help to 
improve the workflow itself even before a publisher is involved. The 
possibility for codechecking papers could be part of a preprint server's
policy, or simply be initiated by interested readers and community members.
% TODO add a sentence about the possible effect of just asking for
% code/data early on in the submission process, see https://twitter.com/tsuyomiyakawa/status/1230672758163402752?s=09)

@@

The topic of blinding is broadly discussed, not the least in the push towards
open peer review as part of the Open Science movement 
(cf. \cite{ross-hellauer_guidelines_2019}).
Without taking a strong stand on this discussion, the general motivation 
for more  transparency and reproducibility does indeed favour a more open 
review process.
However, anonymity can also serve to protect individuals 
\ref{tennant_limitations_2020}, e.g.,  juniour scientists, whereas it might 
be acceptable to require revealing of the names of tenured researchers.
The possible negative effects of a non-anonymous review are greatly reduced
if a CODECHECK is not relevant for the decision to accept or reject, but that
is of course not desirable. Instead, we argue that a CODECHECK is technical
process that should in any case be fixable and not a question of opinion or
faulty approach. It is possible for a nonsense workflow to receive a 
CODECHECK certificate.
Furthermore, the level of documentation that is needed for third parties
to reproduce a workflow is extremely hard to get right. Too often, this 
uncertainty leads to researchers giving up and not documenting at all.

@@

As a prerequisite, we see a CODECHECK as part of a typical scholarly
communication process. While more radical ideas to change the paradigms
how researchers share their work exist\footnote{For example Octopus, 
\url{https://science-octopus.org/about}, or Hypergraph
\url{https://www.libscie.org/hypergraph}.}, there is a 
value in an evolutionary approach. Thereby, existing communities can 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
In our first checks, we found that a codecheck may generally take from 2-5
hours, with some outliers on the higher end. This time includes writing and
publishing the certificate, but excludes the time for actual computations.
The efforts include considerable time to communicate about the process, 
especially around who publishes what and when, so that proper 
cross-referencing between paper and certificate is ensured.
When integrated into peer review systems, this handling of the steps should
become much more effective.

@@

\textbf{Handling failure}: we have no established process for the case
that a reproduction fails. This may be the sign of a pre-selection bias
of the examples, as so far all our codechecks succeeded.
In case of a journal adopting CODECHECK for all submissions, the open 
question remains as what to do in this case will have to be answered.
At the start, we doubt publicly reporting failures (i.e., the code 
would not run) will increase overall reproducibility and recommend not
to share the outcome publicly but to share the latest draft of the 
certific with the authors.
One must be careful as to the impact of acceptance by authors and the
concerns of volunteers this approach, or a contrary one, has.
Private feedback for failures and public acknowledgement/encouragement
for successes seems to be the way to go.
\cite{Rosenthal2016b} discuss incentives for different actors around the
implementation of reproducibility. We see CODECHECK as one mean for
organisations to invest in reproducibility by creating incentives
until reproducible computations become the norm.

@@

\textbf{Who's got time for more peer review?} Agree.
However, there are differences to peer review.
First, the technical nature of a CODECHECK can provide a more 
clearly defined scope and expectations compared to conventional 
peer review. The authors are told what to provide and the codechecker
can be told when to stop, for example.
Second, the specific skill set allows to introduce different groups
into the peer review process. ECRs might be attracted to working with
groups to learn more about recent methods and reproducibility practices.
Research software experts (Research Software Engineers, RSE) who might
not regularly be involved in writing or reviewing papers could be 
interested to increase their connection with scholarly practices.
An extra codechecker may simplify the matchmaking an editor has to do
when identifying suitable reviewers for a submission, as technical and
topical expertise can be provided by different people, who must not even
be traditional peer reviewers. In fact, the codechecker could even be
a software professional employed by a publisher.
Third, recall that we expect CODECHECKSs to be not anonymous and always
publicly deposited, which is common practice for some journals, but not
broadly practiced. The possibility to ask for clarification or more 
material/information can greatly reduce the digging required and thereby
the time spend.
We also found that the focus on the mechanics of a
workflow and the ability to interact with the authors can make
reproductions very educational. It's a different role and as such could
be a welcome alternation/diversion for researchers to give back their 
time to the community.

>>>>

\begin{itemize}
\item Check with Giuliano re: ``code works''. Write to Naomi.
\item Use "execution" insteand of "sharing" in title?
\item Can we introduce "research compendium" as a term?
\item Should we make code and data sharing a prerequisite or leave that door open?
\item Take a look at the Lancet's statistical review process (areas for improvement by authors) \url{https://www.thelancet.com/journals/lancet/article/PII0140-6736(92)90409-V/fulltext}
\item How to relate to "reproducibility does not matter but trust does", see \url{https://gmbecker.github.io/MayInstituteKeynote2019/outline.html}
\item Should we ask the people currently listed in the acknowledgements?
\item revisit \url{https://github.com/codecheckers/discussion}
\item \url{https://hdsr.mitpress.mit.edu/pub/f0obb31j/release/1} and other articles form the HDSR special issue
\item \url{https://av.tib.eu/media/42484} > "Code will make systematic mistakes, not random mistakes", "code is more reliable (and auditable) than a human"
\item \url{https://doi.org/10.1016/j.isci.2020.101698} - "Rigor and Transparency Index may serve as a proxy for methodological quality" - CODECHECK doesn't evaluate that, could be used for distinction
\item Make code Feedback and leave better than found a principle, not an option? Just checking the code does not give enough credit (?), but improving it certainly does, and good peer review improves (watch out to not do full software review though)
\end{itemize}

@@

The point in time when a CODECHECK happens greatly depends on the 
intentions and the ascribed
importance. The earlier a CODECHECK happens in the process, the more the
results can play a role in the most important decision in scholarly 
communication: is a paper published, sent back for revisions, or rejected.
The later in a review process the check happens, the less challenging is it
to allow a bidirectional communication between author and codechecker, e.g.,
because the author might already be notified of the acceptance and may more
freely share materials online.
Even earlier checks, i.e., a CODECHECK of a preprint, may even help to 
improve the workflow itself even before a publisher is involved. The 
possibility for codechecking papers could be part of a preprint server's
policy, or simply be initiated by interested readers and community members.
% TODO add a sentence about the possible effect of just asking for
% code/data early on in the submission process, see https://twitter.com/tsuyomiyakawa/status/1230672758163402752?s=09)

@@

The topic of blinding is broadly discussed, not the least in the push towards
open peer review as part of the Open Science movement 
(cf. \cite{ross-hellauer_guidelines_2019}).
Without taking a strong stand on this discussion, the general motivation 
for more  transparency and reproducibility does indeed favour a more open 
review process.
However, anonymity can also serve to protect individuals 
\cite{tennant_limitations_2020}, e.g.,  juniour scientists, whereas it might 
be acceptable to require revealing of the names of tenured researchers.
The possible negative effects of a non-anonymous review are greatly reduced
if a CODECHECK is not relevant for the decision to accept or reject, but that
is of course not desirable. Instead, we argue that a CODECHECK is technical
process that should in any case be fixable and not a question of opinion or
faulty approach. It is technically possible for a nonsense workflow to receive
a CODECHECK certificate.
If passing a CODECHECK becomes mandatory, the full transparency may have to be 
revisited as the relations between authors and codecheckers fall under the
same social and acommunity challenges as non-anonymous peer review
(cf.~\cite{everythinghertz123}).
% previous sentenc inspired by https://osf.io/9cftx/
Furthermore, the level of documentation that is needed for third parties
to reproduce a workflow is extremely hard to get right. Too often, this 
uncertainty leads to researchers giving up and not documenting at all.

@@

As a prerequisite, we see a CODECHECK as part of a typical scholarly
communication process. While more radical ideas to change the paradigms
how researchers share their work exist\footnote{For example Octopus, 
\url{https://science-octopus.org/about}, or Hypergraph
\url{https://www.libscie.org/hypergraph}.},
% mimosa is another platform idea, but no real prototype AFAICT:
% https://projects.invisionapp.com/share/E9Z3RKE7W3P#/screens/436978924
% https://docs.google.com/presentation/d/1V_K8hghgnvGfEtW7TdwtowTveZyy4Qs8rMemNyHx_Dg/edit#slide=id.p
there is a 
value in an evolutionary approach. Thereby, existing communities can 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
In our first checks, we found that a codecheck may generally take from 2-5
hours, with some outliers on the higher end. This time includes writing and
publishing the certificate, but excludes the time for actual computations.
The efforts include considerable time to communicate about the process, 
especially around who publishes what and when, so that proper 
cross-referencing between paper and certificate is ensured.
When integrated into peer review systems, this handling of the steps should
become much more effective.

@@

\textbf{Handling failure}: We have no established process for the case
that a reproduction fails. This may be the sign of a pre-selection bias
of the examples, as so far all our codechecks succeeded.
In case of a journal adopting CODECHECK for all submissions, the open 
question remains as what to do in this case will have to be answered.
At the start, we doubt publicly reporting failures (i.e., the code 
would not run) will increase overall reproducibility and recommend not
to share the outcome publicly but to share the latest draft of the 
certificate with the authors.
A negative comment in a CODECHECK certificate or a failed check does
not necessarily mean the paper or the science is bad
(cf. discussion on negative comments in \cite{everythinghertz123}).
Therefore, one must be careful as to the impact of acceptance by authors 
and the concerns of volunteers this approach, or a contrary one, has.
Private feedback for failures and public acknowledgement/encouragement
for successes seems to be the way to go.
% eLife's new model (see https://osf.io/9cftx/) will publish reports evantually, e.g., after preprint has been accepted elsewhere
\cite{Rosenthal2016b} discuss incentives for different actors around the
implementation of reproducibility. We see CODECHECK as one mean for
organisations to invest in reproducibility by creating incentives
until reproducible computations become the norm.

@@

\textbf{Who's got time for more peer review?} Agree.
Ignoring the complex community contract behind volunteered peer review,
the time spent with CODECHECKs likely reduces the individual's time to work
as a reviewer.
However, there are differences to peer review.
First, the technical nature of a CODECHECK can provide a more 
clearly defined scope and expectations compared to conventional 
peer review. The authors are told what to provide and the codechecker
can be told when to stop, for example.
Second, the specific skill set allows to introduce different groups
into the peer review process. ECRs might be attracted to working with
groups to learn more about recent methods and reproducibility practices.
Research software experts (Research Software Engineers, RSE) who might
not regularly be involved in writing or reviewing papers could be 
interested to increase their connection with scholarly practices.
An extra codechecker may simplify the matchmaking an editor has to do
when identifying suitable reviewers for a submission, as technical and
topical expertise can be provided by different people, who must not even
be traditional peer reviewers. In fact, the codechecker could even be
a software professional employed by a publisher.
Third, recall that we expect CODECHECKSs to be not anonymous and always
publicly deposited, which is common practice for some journals, but not
broadly practiced. The possibility to ask for clarification or more 
material/information can greatly reduce the digging required and thereby
the time spend.
We also found that the focus on the mechanics of a
workflow and the ability to interact with the authors can make
reproductions very educational. It's a different role and as such could
be a welcome alternation/diversion for researchers to give back their 
time to the community.
